GPU-friendly Stroke Expansion

### 2024-04-28 paper.tex
GPU encoding perf section

Like these other approaches, we also use an intermediate curve, but our choice is an Euler spiral. In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpoliation -- but differs in others. It has no difficulty modeling an inflection point. Further, its parallel curve has a particularly simple mathematical definition and clean behavior regarding cusps.

@@

The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segemnts, offset by the half linewidth. The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky. Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.

@@

\begin{figure}
    \includegraphics{espc}
    \caption{Integral of subdivision density for Euler spiral parallel curve, and its approximation}
    \Description{A graph of the integral of subdivision density. It is a wriggly diagonal line, with flat parts at -1 and 1. The approximation is nearly indistinguishable from the exact value.}
    \label{fig:espc}
\end{figure}

@@

\begin{figure}
    \includegraphics[scale=0.75]{cubic_err_measured} \includegraphics[scale=0.75]{cubic_err_approx}
    \caption{Comparision of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
    \Description{TODO}
    \label{fig:cubic_err}
\end{figure}

@@

Our compute shader is arranged as a 1-dimensional grid, parallelized over input path segments. We chose a workgroup size of 256 as it works well over a wide range of GPUs. However, this choice of workgroup size is not particularly important for our algorithm as it does not require any cross-thread communication. Each thread in the global grid is assigned to process a single path segment. We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch. This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths. We submit the kernel to the GPU as a single dispatch with as many workgroups as needed to handle the entire input.

@@

\begin{table}
    \caption{Comparison of input and output segment counts in the more intensive GPU test cases. \emph{Segments} are input path segments. \emph{Lines} and \emph{Arcs} are the output primitives.}
    \begin{tabular}{|l|c|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Input Segments} & \textbf{Lines} & \textbf{Arcs} \\
    \hline
        waves.svg & 13,308 & 406,059 & 111,965 \\
        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
        long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
        long dash (round) & 503,304 & 2,013,216 & 1,060,476 \\
    \hline
    \end{tabular}
    \label{table:bump-counts}
\end{table}

@@

Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
TODO: brief assessment.

% TODO: line soup is bump allocated. talk about allocation and estimation.

>>>>

Like these other approaches, we also use an intermediate curve, but our choice is an Euler spiral. In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpolation -- but differs in others. It has no difficulty modeling an inflection point. Further, its parallel curve has a particularly simple mathematical definition and clean behavior regarding cusps.

@@

The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth. The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky. Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.

@@

\begin{figure}
    \includegraphics[scale=0.8]{espc}
    \caption{Integral of subdivision density for Euler spiral parallel curve, and its approximation}
    \Description{A graph of the integral of subdivision density. It is a wriggly diagonal line, with flat parts at -1 and 1. The approximation is nearly indistinguishable from the exact value.}
    \label{fig:espc}
\end{figure}

@@

\begin{figure}
    \includegraphics[scale=0.75]{cubic_err_measured} \includegraphics[scale=0.75]{cubic_err_approx}
    \caption{Comparison of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
    \Description{TODO}
    \label{fig:cubic_err}
\end{figure}

@@

Our compute shader is arranged as a 1-dimensional grid, parallelized over input path segments. We chose a workgroup size of 256 as it works well over a wide range of GPUs. However, this choice of workgroup size is not particularly important for our algorithm as it does not require any cross-thread communication. Each thread in the global grid is assigned to process a single path segment. We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondence to a global thread ID in the dispatch. This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths. We submit the kernel to the GPU as a single dispatch with as many workgroups as needed to handle the entire input.

@@

\begin{table}
    \caption{Comparison of input and output segment counts in the more intensive GPU test cases. \emph{Segments} are input path segments. \emph{Lines} and \emph{Arcs} are the output primitives.}
    \begin{tabular}{|l|c|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Input Segments} & \textbf{Lines} & \textbf{Arcs} \\
    \hline
        waves.svg & 13,308 & 406,059 & 111,965 \\
        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
        long path (butt) & 503,304 & 1,060,476 & 1,060,476 \\
        long path (round) & 503,304 & 2,013,216 & 1,060,476 \\
    \hline
    \end{tabular}
    \label{table:bump-counts}
\end{table}

@@

Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets. The entire Nehab\ref{2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit. Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of rendering \emph{waves} in 3 ms. We observed that performance of our kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested with. We also observed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.

\subsection{GPU: Encoding} \label{subsection:encoding-results}

Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size. The prefix sums scale better at large input sizes compared to our stroke expansion kernel. This can be attributed to the lack of expensive computations and divergence the tag monoid kernel.

Our GPU implementation targets graphics APIs that do not support dynamic memory allocations in shader code. This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimates}). For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth. We observed that this increases our CPU encoding time by 2-3x on the larger scenes.

\begin{figure}
    \includegraphics[scale=0.69]{tag_monoid_gpu_timings}
    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090. The figure also shows the execution time of the stroke expansion and flattening kernel and compares their scaling with input size}
    \Description{Mean GPU execution time of the tag monoid kernel}
    \label{fig:tag-monoid-gpu-timings}
\end{figure}

\begin{table}
    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes. All measurements were computed on an Apple M1 Max CPU.}
    \begin{tabular}{|l|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding (ms)} & \textbf{w/ Estimation (ms)} \\
    \hline
        waves.svg & 0.244 & 0.460 \\
        mmark-35k & 2.45 & 6.7 \\
        mmark-60k & 4.17 & 11.8 \\
        long path & 4.14 & 6.28 \\
    \hline
    \end{tabular}
    \label{table:bump-estimates}
\end{table}
